---
title: "Interaction_Effects"
format: html
editor: visual
---

# Interaction Effects

## Data Cleaning

```{r}
library(tidyverse)

setwd("/Users/katewest/Desktop/330 final project")

proj_data <- read_delim("Spotify_Dataset_V3.csv", delim = ";")

spotify_clean <- proj_data %>%
  select(
    Title,
    Rank,
    Artists,
    Danceability,
    Energy,
    Loudness,
    Speechiness,
    Acousticness,
    Instrumentalness,
    Valence
  ) %>%
  mutate(
    multi_artist = ifelse(str_detect(Artists, ","), 1, 0),
    rank_score = 201 - Rank
  ) %>%
  group_by(Title) %>%
  mutate(
    days_in_top200 = n()
  ) %>%
  ungroup()

spotify_data <- spotify_clean %>%
  group_by(Title, Artists) %>%
  summarise(
    avg_rank_score = mean(rank_score),
    days_in_top200 = n(),
    multi_artist = first(multi_artist),
    Danceability = first(Danceability),
    Energy = first(Energy),
    Loudness = first(Loudness),
    Speechiness = first(Speechiness),
    Acousticness = first(Acousticness),
    Instrumentalness = first(Instrumentalness),
    Valence = first(Valence),
    .groups = "drop"
  )

head(spotify_data)
```

## Baseline Regression Models (No Interactions)

```{r}
model_top200_main <- lm(
  days_in_top200 ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist,
  data = spotify_data
)

model_rank_main <- lm(
  avg_rank_score ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist,
  data = spotify_data
)

summary(model_top200_main)
summary(model_rank_main)
```

## Add Interactions

```{r}
#Danceability and Energy for Top 200
model_top200_int1 <- lm(
  days_in_top200 ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Danceability:Energy,
  data = spotify_data
)

summary(model_top200_int1)
anova(model_top200_main, model_top200_int1)

#NOT at all even close to significant LOL

#Danceabiility and Energy for Rank
model_rank_int1 <- lm(
  avg_rank_score ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Danceability:Energy,
  data = spotify_data
)

summary(model_rank_int1)
anova(model_rank_main, model_rank_int1)
#Almost significant tbh


# Danceability and Valence for Top 200
model_top200_int2 <- lm(
  days_in_top200 ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Danceability:Valence,
  data = spotify_data
)

summary(model_top200_int2)
anova(model_top200_main, model_top200_int2)

#Very close to significant, right on .05

#Danceability and Valence for Rank
model_rank_int2 <- lm(avg_rank_score ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Danceability:Valence,
  data = spotify_data
)

summary(model_rank_int2)
anova(model_rank_main, model_rank_int2)

#Nah not really


#Energy and Loudness for Top 200
model_top200_int3 <- lm(
  days_in_top200 ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Energy:Loudness,
  data = spotify_data
)

summary(model_top200_int3)
anova(model_top200_main, model_top200_int3)

#Absolutely NOT significant

#Energy and Loudness for Rank
model_rank_int3 <- lm(
  avg_rank_score ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Energy:Loudness,
  data = spotify_data
)

summary(model_rank_int3)
anova(model_rank_main, model_rank_int3)

#Very significant!!!!! For predicting rank
```

-   The **Danceability × Energy** interaction was **not significant** for predicting days in the Top 200, and showed only weak evidence for predicting average rank score.
    The **Danceability × Valence** interaction showed **borderline evidence** for predicting days in the Top 200 but was **not significant** for predicting rank.
    The **Energy × Loudness** interaction was **not significant** for predicting days in the Top 200
    However, the **Energy × Loudness interaction was statistically significant for predicting average rank score**, indicating that the relationship between energy and rank depends on the level of loudness.

## Several Interactions Together

```{r}
model_top200_allint <- lm(
  days_in_top200 ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Danceability:Energy +
    Danceability:Valence +
    Energy:Loudness +
    Speechiness:Acousticness,
  data = spotify_data
)

summary(model_top200_allint)
anova(model_top200_main, model_top200_allint)


model_rank_allint <- lm(
  avg_rank_score ~ Danceability + Energy + Loudness +
    Speechiness + Acousticness + Instrumentalness +
    Valence + multi_artist +
    Danceability:Energy +
    Danceability:Valence +
    Energy:Loudness +
    Speechiness:Acousticness,
  data = spotify_data
)

summary(model_rank_allint)
anova(model_rank_main, model_rank_allint)
```

## Side Quest: Rank and Days in Top 200

```{r}
ggplot(spotify_data, aes(x = avg_rank_score, y = days_in_top200)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Rank and Days in Top 200",
    x = "Average Rank Score (Higher = Better)",
    y = "Days in Top 200"
  ) +
  theme_minimal()


cor(spotify_data$avg_rank_score, spotify_data$days_in_top200)

model_relationship <- lm(
  days_in_top200 ~ avg_rank_score,
  data = spotify_data
)

summary(model_relationship)

#Flipped
model_relationship2 <- lm(
  avg_rank_score ~ days_in_top200,
  data = spotify_data
)

summary(model_relationship2)


#with controls!
model_relationship_controls <- lm(
  days_in_top200 ~ avg_rank_score + Danceability + Energy +
    Loudness + Speechiness + Acousticness +
    Instrumentalness + Valence + multi_artist,
  data = spotify_data
)

summary(model_relationship_controls)
```

# Conclusion

Most interaction effects were not statistically significant, meaning that audio features largely act independently. However, the Energy × Loudness interaction significantly improved prediction of average rank score, meaning that the effect of energy on rank changes depending on loudness. Additionally, Speechiness × Acousticness was significant for predicting days in the Top 200. A moderate positive relationship (r ≈ 0.31) was also observed between Rank and Days in Top 200, indicating that higher-ranked songs tend to stay on the chart longer.
